---
title: "R Notebook"
output: html_notebook
---


```{r}
x <- 10 # create an object called x
```

```{r}
x + 5
```

```{r}
function_name(arg1 = , arg2 = , arg3, ....)
```

```{r}
mean(c(5,2,5,2,10,25,NA), na.rm = T)
```
# This is the title

## Load in my data

### Load in temperature data

### Load in precipitation data

## Analyze my weather data

Often we want to **bold our font** and sometimes we want *to make it italic* another way to italic is _this_

`make it look like code in my text`


x^the^ 

CO~2~


1. Item one  
2. Item two  
    i) sub item  
    ii) sub item  


Here is my text and I want a line break  

<br/>

Here is my next line


# Basic operations in R

```{r}
2 + 2 # add
10 -5 # subtraction

10*10 # multiply

3^4 # exponents

10/2.5 # division


```
```{r}
 51 %% 7

25 %% 12  # modulus
```

```{r}
4 > 2 # greater than 
```

TRUE
T
1

FALSE
F
0


```{r}
4 < 2 # less than

4 >= 2 

4 <= 2
```

```{r}
10 == 10

10 != 5
```


```{r}
name_student_01 <- "Mason"
name_student_02 <- "Stahl"

studentname3 <- "Mike"
```

```{r}
name_student_02
```


```{r}
ls()
```

```{r}

```





```{r}
df_char <- "a"  # character

df_numeric <- 2.25 # numeric

df_int <-  as.integer(5)

df_logical <- T 

```



```{r}
1/0
```

```{r}
0/0
```





## Data structures 

- vector
- lists
- matrices
- data frames
- tables


### Vector
```{r}
majors_vec <- c("Environmental Science","Geology","Chemistry")
majors_vec

```

```{r}
num_vec <- c(1, 10, 5034.253, -1.045)
```



```{r}
logical_vec <- c(T, F, F, F, T)

logical_vec
```

```{r}
length(logical_vec)
```

```{r}
class(num_vec)
```

































