---
title: "Contour maps"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: no
---

<br/>



```{r echo = F}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = F)
```


```{r}
library(tidyverse)
library(rayshader)
library(stars)
library(tmap)
library(plot3D)
```




```{r}
surface <- reshape2::melt(volcano)

plot_volcano_3D <- ggplot(surface, aes(Var1, Var2, z = value)) +
  geom_raster(aes(Var1, Var2, fill = value)) +
  scale_fill_gradientn(colours = terrain.colors(20)) 
  

```

```{r eval = F}
rayshader::plot_gg(plot_volcano_3D , max_error = 0.005,
                   multicore=TRUE,
                   width=7,height=7,scale=600)

rgl::rglwidget()

rgl::rgl.close()
```



```{r eval = F}
ggplot(surface, aes(Var1, Var2, z = value)) +
  geom_raster(aes(Var1, Var2, fill = value)) +
  scale_fill_gradientn(colours = terrain.colors(20)) +
  geom_contour(color = "black", size = 0.1)
```


## Three dimensional object

Below is a 3D rendering of a volcano.  Any point on a surface has three spatial dimensions associated with it, namely it's **x** (longitude), **y** (latitude), and **z** (elevation).  

```{r eval = T}

volcano %>%
  sphere_shade(texture = "desert") %>%
  add_shadow(ray_shade(volcano,zscale=3),0.3) %>%
  #add_overlay(generate_contour_overlay(volcano_elev, linewidth = 0.5))  %>%
  plot_3d(volcano, zscale=3, fov=30)

rgl::rglwidget()

rgl::rgl.close()
```


```{r eval = F}
volcano %>%
  sphere_shade(texture = "imhof4") %>%
  add_shadow(ray_shade(volcano, zscale = 3), 0.5) %>%
  add_shadow(ambient_shade(volcano), 0) %>%
  add_overlay(generate_contour_overlay(volcano, linewidth = 0.2))  %>%
  plot_3d(volcano, zscale = 1, fov = 0, theta = -30, zoom = 0.75, phi = 45, windowsize = c(1000, 800))
Sys.sleep(0.2)
#render_depth(focus = 0.6, focallength = 10, clear = TRUE)
render_snapshot(clear=TRUE)

rgl::rglwidget()

rgl::rgl.close()
```




<br/>

## A volcano represented in two dimensions
It is often useful/practical to represent three spatial dimensions on a flat (2 dimensional) surface (e.g., a sheet of paper).  We can do this in a number of ways, one of which is to represent the third spatial dimension (elevation) as color.  In the 2D image below we represent the x,y spatial dimensions on the x,y spatial dimensions of the image and we represent the elevation using color.  

```{r}
image2D(volcano, contour = F, col = hcl.colors(100, "terrain") ) 
```

<br/>

We can also represent the third dimension (elevation) using contours.  A given contour connects points with the same value - in this case a given contour represents a particular elevation.  Every point on a given contour has the exact same elevation.  

```{r}
image2D(volcano, contour = TRUE, col = hcl.colors(100, "terrain") )
```


<br/>


Below is the same data represented as a surface in three dimensions with the contour map (shown above the colored surface) projected down onto the surface.  


```{r}
persp3D(z = volcano, contour = list(side = c("zmax", "z")), zlim= c(90, 300),
        col = hcl.colors(100, "terrain"),
        phi = 30, theta = 20, d = 10, box = FALSE)
```


<br/>

***
| | |
|--|--|
| _Mason Stahl (Union College)_ |  |
