---
title: "Exploring Environmental Data (ENS-215)"
date: "Winter 2020"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
---

<!-- Global site tag (gtag.js) - Google Analytics -->
<script async src="https://www.googletagmanager.com/gtag/js?id=UA-137229217-1"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'UA-137229217-1');
</script>

***
|  [ENS-215](https://stahlm.github.io/ENS_215/Winter_2020/Home.html) | [Homework](https://stahlm.github.io/ENS_215/Winter_2020/HW.html) | [Syllabus](https://stahlm.github.io/ENS_215/Winter_2020/Syllabus.html) |
|---|------------|-----------|


## Lectures 

### Week 5
+ Mon (3-Feb) -- **Presentation quality graphics**
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_02_03.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_02_03.Rmd), [(Slides)](https://stahlm.github.io/ENS_215/Documents/Data_vis_slides.pdf), [Article by Kelleher and Wagener (2011)](https://github.com/stahlm/stahlm.github.io/raw/master/ENS_215/Documents/ems2011%20kelleher%20and%20wagener.pdf) 
```{r eval = F, echo = F}
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_02_03.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_02_03.Rmd), [(Slides)](https://stahlm.github.io/ENS_215/Documents/Data_vis_slides.pdf), [Article by Kelleher and Wagener (2011)](https://github.com/stahlm/stahlm.github.io/raw/master/ENS_215/Documents/ems2011%20kelleher%20and%20wagener.pdf) 
```


***

### Week 4
+ Mon (27-Jan) -- **Intro to data visualization**
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_27_Intro_data_viz.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_27_Intro_data_viz.Rmd), [Slides](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_27_slides.html) 

+ Wed (29-Jan) -- **Intro to data viz (continued)**
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_29_Intro_data_viz.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_29_Intro_data_viz.Rmd)

+ Fri (31-Jan) -- **Data visualization (additional topics)**
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_31_Data_viz.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_31_Data_viz.Rmd) 


```{r echo = F, eval = F}
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_31_Data_viz.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_31_Data_viz.Rmd) 

```


***

### Week 3
+ Mon (20-Jan) -- **Data wrangling continued** 
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_20_Data_manipulation_cont.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_20_Data_manipulation_cont.Rmd)  

+ Wed (22-Jan) -- **Data wrangling `group_by` and `summarize()`**
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_22.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_22.Rmd) 

+ Fri (24-Jan) -- **Intro to strings** 
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_24_strings.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_24_strings.Rmd)  



***

### Week 2
+ Mon (13-Jan) -- **Relational and logical operators, if/else/else if** 
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_13_R_basics_cont.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_13_R_basics_cont.Rmd)

+ Wed (15-Jan) -- **Control flow (for and while loops)**
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_15_R_basic_loops.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_15_R_basic_loops.Rmd)

+ Fri (17-Jan) --**Data wrangling intro**
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_17_Data_manipulation.html)  [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_17_Data_manipulation.Rmd)




***

### Week 1
+ Mon (6-Jan) -- **Course introduction** [(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_06_L1.html) [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_06_L1.Rmd), [Slides](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/Intro_Survey.html)  

+ Wed (8-Jan) -- **Diving into R** [(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_08_L1_diving_in.html) [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_08_L1_diving_in.Rmd), [Slides](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_08_Slides.html), [Cheatsheets](https://stahlm.github.io/ENS_215/Resources/Cheatsheets_all.pdf)

+ Fri (10-Jan) -- **R Programming and Markdown Basics**
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Lectures/2020_01_10_R_basics.html) [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Lectures/2020_01_10_R_basics.Rmd)


***
<br/>

## Labs
+ Lab 1 -- **Computer setup and logistics**

+ Lab 2 -- **Intro and EPA Fuel Economy Data** 
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Labs/Lab_EPA_Fuel_Economy.html) [(Rmd)](https://raw.githubusercontent.com/stahlm/stahlm.github.io/master/ENS_215/Winter_2020/Labs/Lab_EPA_Fuel_Economy.Rmd)  

+ Lab 3 -- **Global Dams and Reservoirs** [(html)](https://stahlm.github.io/ENS_215/Winter_2020/Labs/Lab_Global_Dams.html)  

+ Lab 4 -- **Groundwater Arsenic in Bangladesh** 
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Labs/Lab_Bangladesh_Arsenic.html)

+ Lab 5 -- **Final project kick-off** 


```{r echo=F,eval=F}
[(html)](https://stahlm.github.io/ENS_215/Winter_2020/Labs/Lab_Final_Project.html)
```


<br/>

***
| | |
|--|--|
| _Mason Stahl (2020)_ |  |
| _Union College_ | |